---
title: "Lab 05 - Data Wrangling"
output: 
  github_document: default
---

# Learning goals

- Use the `merge()` function to join two datasets.
- Deal with missings and impute data.
- Identify relevant observations using `quantile()`.
- Practice your GitHub skills.

# Lab description

For this lab we will be dealing with the meteorological dataset `met`. In this case, we will use `data.table` to answer some questions regarding the `met` dataset, while at the same time practice your Git+GitHub skills for this project.

This markdown document should be rendered using `github_document` document.

# Part 1: Setup a Git project and the GitHub repository

1. Go to wherever you are planning to store the data on your computer, and create a folder for this project

2. In that folder, save [this template](https://github.com/JSC370/jsc370-2023/blob/main/labs/lab05/lab05-wrangling-gam.Rmd) as "README.Rmd". This will be the markdown file where all the magic will happen.

3. Go to your GitHub account and create a new repository of the same name that your local folder has, e.g., "JSC370-labs".

4. Initialize the Git project, add the "README.Rmd" file, and make your first commit.

5. Add the repo you just created on GitHub.com to the list of remotes, and push your commit to origin while setting the upstream.

Most of the steps can be done using command line:

```sh
# Step 1
cd ~/Documents
mkdir JSC370-labs
cd JSC370-labs

# Step 2
wget https://raw.githubusercontent.com/JSC370/jsc370-2023/main/labs/lab05/lab05-wrangling-gam.Rmd
mv lab05-wrangling-gam.Rmd README.Rmd
# if wget is not available,
curl https://raw.githubusercontent.com/JSC370/jsc370-2023/main/labs/lab05/lab05-wrangling-gam.Rmd --output README.Rmd

# Step 3
# Happens on github

# Step 4
git init
git add README.Rmd
git commit -m "First commit"

# Step 5
git remote add origin git@github.com:[username]/JSC370-labs
git push -u origin master
```

You can also complete the steps in R (replace with your paths/username when needed)

```r
# Step 1
setwd("~/Documents")
dir.create("JSC370-labs")
setwd("JSC370-labs")

# Step 2
download.file(
  "https://raw.githubusercontent.com/JSC370/jsc370-2023/main/labs/lab05/lab05-wrangling-gam.Rmd",
  destfile = "README.Rmd"
  )

# Step 3: Happens on Github

# Step 4
system("git init && git add README.Rmd")
system('git commit -m "First commit"')

# Step 5
system("git remote add origin git@github.com:[username]/JSC370-labs")
system("git push -u origin master")
```

Once you are done setting up the project, you can now start working with the MET data.

## Setup in R

1. Load the `data.table` (and the `dtplyr` and `dplyr` packages if you plan to work with those).

2. Load the met data from https://github.com/JSC370/jsc370-2023/blob/main/labs/lab03/met_all.gz or  (Use https://raw.githubusercontent.com/JSC370/jsc370-2023/main/labs/lab03/met_all.gz to download programmatically), and also the station data. For the latter, you can use the code we used during lecture to pre-process the stations data:

```{r setup}
library(data.table)
library(dtplyr) # translator between dplyr (tidyverse) and data.table
library(dplyr)
library(ggplot2)
library(leaflet)
```

```{r stations-data, eval = TRUE}
# Download the data
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]
```


```{r}
# read met dataset
if (!file.exists("met_all.gz")) {
 download.file(
      url = "https://raw.githubusercontent.com/JSC370/jsc370-2023/main/labs/lab03/met_all.gz",
      destfile = "met_all.gz",
      method   = "libcurl",
      timeout  = 60
      ) 
}
met <- data.table::fread("met_all.gz")
```

3. Merge the data as we did during the lecture.

```{r}
# data.table version
met <- merge(
  x = met,
  y = stations,
  all.x = TRUE, all.y = FALSE,
  by.x = "USAFID", by.y = "USAF"
)
```

```{r eval=FALSE}
# same operation in tidyverse (don't run)
met <- left_join(
  met, stations, by = c("USAFID" = "USAF")
)
```

```{r}
# lazy version for faster tidyverse operation
met_lz <- lazy_dt(met, immutable = FALSE)
```

## Question 1: Representative station for the US

Across all weather stations, what is the median station in terms of temperature, wind speed, and atmospheric pressure? Look for the three weather stations that best represent continental US using the `quantile()` function. Do these three coincide?

```{r}
# average for each station
met_avg_lz <- met_lz |>
  group_by(USAFID, STATE) |>  # keep state info for Q2 & Q3
  summarise(
    across(
      c(temp, wind.sp, atm.press, lon, lat),  # lon and lat for Q3
      function(x) mean(x, na.rm = TRUE)
    ), .groups = "drop"
  )
# find medians of temp, wind.sp, atm.press
met_med_lz <- met_avg_lz |>
  summarise(across(
    3:5,
    function(x) quantile(x, probs = .5, na.rm = TRUE)
  ))
# temperature
temp_us_id <- met_avg_lz |>
  mutate(
    d = abs(temp - met_med_lz |> pull(temp))
    ) |>
  arrange(d) |>
  slice(1) |>
  pull(USAFID)
# wind speed
wsp_us_id <- met_avg_lz |>
  mutate(
    d = abs(wind.sp - met_med_lz |> pull(wind.sp))
    ) |>
  arrange(d) |>
  slice(1) |>
  pull(USAFID)
# atm speed
atm_us_id <- met_avg_lz |>
  mutate(
    d = abs(atm.press - met_med_lz |> pull(atm.press))
    ) |>
  arrange(d) |>
  slice(1) |>
  pull(USAFID)
# Display with the geolocation info
met_lz |>
  group_by(USAFID) |>
  # smome stations have multiple values
  summarise(lon = mean(lon), lat = mean(lat)) |> 
  filter(USAFID %in% c(temp_us_id, wsp_us_id, atm_us_id)) 
```

```{r}
# for "pretty" printing
met_lz |>
  group_by(USAFID) |>
  # smome stations have multiple values
  summarise(lon = mean(lon), lat = mean(lat)) |> 
  filter(USAFID %in% c(temp_us_id, wsp_us_id, atm_us_id)) |>
  mutate("Variable" = c("Temperature", "Wind speed", "Atmospheric pressure")) |>
  select(Variable, USAFID, lon, lat) |>
  as_tibble()
```


```{r eval=FALSE}
# data.table version
# mean per station
station_avg <- met[ , .(
  temp      = mean(temp, na.rm = TRUE),
  wind.sp   = mean(wind.sp, na.rm = TRUE),
  atm.press = mean(atm.press, na.rm = TRUE),
  lat       = mean(lat),
  lon       = mean(lon)
), by = .(USAFID, STATE)] # keep state info for Q2 & Q3
# medain across the US
usa_median <- station_avg[ , .(
  temp_50      = quantile(temp, probs = .5, na.rm = TRUE),
  wind.sp_50   = quantile(wind.sp, probs = .5, na.rm = TRUE),
  atm.press_50 = quantile(atm.press, probs = .5, na.rm = TRUE)
)]
# find the stations closest to the median values
station_avg[ , d_temp := abs(temp - usa_median$temp_50)]
station_avg[ , d_wind.sp   := abs(wind.sp - usa_median$wind.sp_50)]
station_avg[ , d_atm.press := abs(atm.press - usa_median$atm.press_50)]
# temperature
station_avg[order(d_temp), .(USAFID, STATE, temp)][1]
# wind speed
station_avg[order(d_wind.sp), .(USAFID, STATE, wind.sp)][1]
# atm pressure
station_avg[order(d_atm.press), .(USAFID, STATE, atm.press)][1]
```

Knit the document, commit your changes, and save it on GitHub. Don't forget to add `README.md` to the tree, the first time you render it.

## Question 2: Representative station per state

Just like the previous question, you are asked to identify what is the most representative, the median, station per state. This time, instead of looking at one variable at a time, look at the euclidean distance. If multiple stations show in the median, select the one located at the lowest latitude.

```{r}
# function for computing Euclidean distance
udist <- function(pt1, pt2) {
  sqrt(sum((pt1 - pt2)^2))
}
# example usage
udist(c(1,1), c(3,3))
# example usage
udist(c(1,1,1), c(3,3,3))
```

```{r}
# tidyverse
# compute median per state
state_med_lz <- met_avg_lz |>
  group_by(STATE) |>
  summarise(
    across(2:4, function(x) quantile(x, probs = .5, na.rm = TRUE),
           .names = "{.col}_50"), # name columns with "_50"
    .groups = "drop"
  )
# find the minimum Euclidean distance stations
state_med_stations_lz <- met_avg_lz |>
  left_join(state_med_lz, by = "STATE") |>
  collect() |> rowwise() |> # to sum() row wise
  mutate(d = udist(c(temp, wind.sp, atm.press), 
                   c(temp_50, wind.sp_50, atm.press_50))) |>
  group_by(STATE) |>
  slice_min(d) # find the minimum d row
state_med_stations_lz
```

Note that the result includes only 46 states. There are 48 state in continental US.

```{r}
# which states are missing?
state_orig <- met |> distinct(STATE) |> pull(STATE) 
state_46 <- state_med_stations_lz |> pull(STATE)
state_missing <- state_orig[!state_orig %in% state_46]
state_med_lz |>
  filter(STATE %in% state_missing) |>
  as_tibble()
```

All stations in ND and WA are missing atmospheric pressure values. One way of handling the issue would be ignoring any measurement with missing values when computing the Euclidean distance.

```{r}
# function for computing Euclidean distance
udist_handle_missing <- function(pt1, pt2) {
  pt1[is.na(pt1) | is.na(pt2)] <- 0
  pt2[is.na(pt1) | is.na(pt2)] <- 0
  sqrt(sum((pt1 - pt2)^2))
}
```

```{r}
# find the minimum Euclidean distance stations
state_med_stations_lz_handle_missing <- met_avg_lz |>
  left_join(state_med_lz, by = "STATE") |>
  collect() |> rowwise() |> # to sum() row wise
  mutate(d = udist_handle_missing(
    c(temp, wind.sp, atm.press), 
    c(temp_50, wind.sp_50, atm.press_50))) |>
  group_by(STATE) |>
  slice_min(d) # find the minimum d row
state_med_stations_lz_handle_missing
```


```{r eval=FALSE}
# data.table version
# medain across the US
station_avg[ , temp_state_50 := quantile(
  temp, probs = .5, na.rm = TRUE
  ), by = STATE]
station_avg[ , wind.sp_state_50 := quantile(
  wind.sp, probs = .5, na.rm = TRUE
  ), by = STATE]
station_avg[ , atm.press_state_50 := quantile(
  atm.press, probs = .5, na.rm = TRUE
  ), by = STATE]
# find the stations closest to the median values
station_avg[ , metdist := udist_handle_missing(
  c(temp, wind.sp, atm.press),
  c(temp_state_50, wind.sp_state_50, atm.press_state_50)),
  by = 1:nrow(station_avg)] # by 1:nrow() has the same effect as row.wise()
# select the top
station_avg[ , .SD[which.min(metdist)], by = STATE]
```


Knit the doc and save it on GitHub.

## Question 3: In the middle?

For each state, identify what is the station that is closest to the mid-point of the state. Combining these with the stations you identified in the previous question, use `leaflet()` to visualize all ~100 points in the same figure, applying different colors for those identified in this question.

```{r warning=FALSE}
# tidyverse
# compute geospatial centre per state
state_centre_lz <- met_avg_lz |>
  group_by(STATE) |>
  summarise(
    across(c(lat, lon), function(x) quantile(x, probs = .5, na.rm = TRUE),
           .names = "{.col}_50"), # name columns with "_50"
    .groups = "drop"
  )
# find the minimum Euclidean distance stations
state_centre_stations_lz <- met_avg_lz |>
  left_join(state_centre_lz, by = "STATE") |>
  collect() |> rowwise() |> # to sum() row wise
  mutate(d = udist(c(lat, lon), 
                   c(lat_50, lon_50))) |>
  group_by(STATE) |>
  slice_min(d) # find the minimum d row
# map
stations_map <- leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addCircles(
    data = state_centre_stations_lz,
    lat = ~lat, lng = ~lon, popup = "Geospatial mid",
    opacity = 1, fillOpacity = 1, radius = 20, color = "darkgreen"
  ) |>
  addCircles(
    data = state_med_stations_lz_handle_missing,
    lat = ~lat, lng = ~lon, popup = "Climate mid",
    opacity = 1, fillOpacity = 1, radius = 20, color = "purple"
  ) |>
  addLegend(
    colors = c("darkgreen", "purple"),
    labels = c("Geospatially middle stations", "Climate representative stations")
  )
# use mapview::mapshot to convert the map to a static image file and
# display on github (credit: Chan Yu)
mapview::mapshot(stations_map, file = "mapview.png")
stations_map
```

_`data.table` version omitted. See Q3._

Knit the doc and save it on GitHub.

## Question 4: Means of means

Using the `quantile()` function, generate a summary table that shows the number of states included, average temperature, wind-speed, and atmospheric pressure by the variable "average temperature level," which you'll need to create. 

Start by computing the states' average temperature. Use that measurement to classify them according to the following criteria: 

- low: temp < 20
- Mid: temp >= 20 and temp < 25
- High: temp >= 25

```{r}
# tidyverse
met_hml_lz <- met_lz |>
  group_by(STATE) |>
  summarise(
    n_entries = n(),
    n_entries_na = sum(is.na(temp)),
    n_stations = n_distinct(USAFID),
    across(c(temp, wind.sp, atm.press), function(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(
    state_temp_cat = if_else(
      temp < 20, "Low",
      if_else(temp < 25, "Mid", "High")
    )
  )
```

Once you are done with that, you can compute the following:

- Number of entries (records),
- Number of NA entries,
- Number of stations,
- Number of states included, and
- Mean temperature, wind-speed, and atmospheric pressure.

All by the levels described before.

```{r}
met_hml_lz |>
  mutate(
    state_temp_cat = factor(state_temp_cat, levels = c("High", "Mid", "Low")) # order
  ) |>
  group_by(state_temp_cat) |>
  summarise(
    n_state = n(),
    n_stations = sum(n_stations),
    n_records = sum(n_entries),
    n_entries_na = sum(n_entries_na),
    temp = mean(temp, na.rm = TRUE),
    wind.sp = mean(wind.sp, na.rm = TRUE),
    atm.press = mean(atm.press, na.rm = TRUE)
  ) |>
  collect()
```

```{r}
# data.table
met[ , state_temp := mean(temp, na.rm = TRUE), by = STATE]
met[ , state_temp_cat := factor(fifelse(
  state_temp < 20, "Low",
  fifelse(state_temp < 25, "Mid", "High")
), levels = c("High", "Mid", "Low"))]
tab <- met[ , .(
  n_state      = length(unique(STATE)),
  n_stations   = length(unique(USAFID)),
  n_entries    = .N,
  n_entries_na = sum(is.na(temp)),
  temp         = mean(temp, na.rm = TRUE),
  wind.sp      = mean(wind.sp, na.rm = TRUE),
  atm.press    = mean(atm.press, na.rm = TRUE)
), keyby = state_temp_cat] # sort by the factor levels
knitr::kable(tab) # knitr::kable() is works well for both html and pdf outputs
```

Knit the document, commit your changes, and push them to GitHub. 


## Question 5: Advanced Regression

Let's practice running regression models with smooth functions on X. We need the `mgcv` package and `gam()` function to do this.

- using your data with the median values per station, examine the association between median temperature (y) and median wind speed (x). Create a scatterplot of the two variables using ggplot2. Add both a linear regression line and a smooth line.

- fit both a linear model and a spline model (use `gam()` with a cubic regression spline on wind speed). Summarize and plot the results from the models and interpret which model is the best fit and why.

```{r warning=FALSE}
met_med_tbl <- met_lz |>
  group_by(USAFID) |>
  summarise(
    across(c(temp, wind.sp), function(x) quantile(x, probs = .5, na.rm = TRUE))
  ) |>
  collect() # turn into tibble before using ggplot
ggplot(met_med_tbl, aes(x = wind.sp, y = temp)) +
  theme_minimal() +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, col = "darkgreen") +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), col = "purple") +
  labs(x = "Wind speed", y = "Temperature")

```


```{r}
lm_mod <- lm(temp ~ wind.sp, data = met_med_tbl)
summary(lm_mod)
plot(lm_mod)
```


```{r}
library(mgcv)
gam_mod <- gam(temp ~ s(wind.sp, bs = "cr", k = 10), data = met_med_tbl)
summary(gam_mod)
plot(gam_mod)
```



